---
title: "Assignment 5: Data Visualization"
author: "Chrissie Pantoja"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

```{r}
# Load necessary packages
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

# Verify your home directory
home_directory <- here::here()
print(home_directory)

# Define paths to the processed data files
peter_paul_file <- here("Data/Processed_KEY", "NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
niwo_litter_file <- here("Data/Processed_KEY", "NEON_NIWO_Litter_mass_trap_Processed.csv")

# Read in the NTL-LTER data for Peter and Paul Lakes
peter_paul_data <- read_csv(peter_paul_file)

# Read in the Niwot Ridge litter dataset
niwo_litter_data <- read_csv(niwo_litter_file)

# Display the first few rows of each dataset to verify successful loading
head(peter_paul_data)
head(niwo_litter_data)

```


2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}

# Check the structure of the data to see the format of the date columns
str(peter_paul_data$sampledate)
str(niwo_litter_data$collectDate)

# Function to convert date strings to Date format (MM-DD-YYY)
#peter_paul_data$sampledate <- as.Date(peter_paul_data$collectDate, format = "%m/%d/%Y")
#niwo_litter_data$collectDate <- as.Date(niwo_litter_data$collectDate, format = "%m/%d/%Y")

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}

# Custom theme
my_custom_theme <- theme(
 
  # Customize plot background
  plot.background = element_rect(fill = "lightgray", color = "black"),  # light gray background with black    border
  
   # Customize plot title
  plot.title = element_text(face = "bold", size = 14, hjust = 0.5, color = "darkblue"),  # centered, bold,    large font
  
  # Customize axis labels
  axis.title.x = element_text(face = "italic", size = 12, color = "darkblue"),  # italic and colored x-axis     label
  axis.title.y = element_text(face = "italic", size = 12, color = "darkblue"),  # italic and colored y-axis     label
  
  # Customize axis ticks/gridlines
  axis.ticks = element_line(color = "blue"),  # blue axis ticks
  panel.grid.major = element_line(color = "gray80", size = 0.5),  # light gray major gridlines
  panel.grid.minor = element_blank(),  # remove minor gridlines
)

# Set the custom theme as the default
theme_set(my_custom_theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}

# Example plot to demonstrate the theme

example_plot <- peter_paul_data %>% 
       ggplot(aes(x = po4, 
           y = tp_ug, 
           color = lakename)) +
  geom_smooth(method = "lm", se = FALSE) +  # Add a line of best fit using the linear model (lm) method
  labs(
    title = "Total Phosphorus vs. Phosphate in Peter and Paul Lakes",
    x = "Phosphate (ug/L)",
    y = "Total Phosphorus (ug/L)",
    color = "Lake Name"
  ) +
  # Adjust axes to hide extreme values (use xlim() and ylim() based on data distribution)
  xlim(0, 50) +  # Set the limits for the x-axis based on reasonable phosphate values
  ylim(0, 100)   # Set the limits for the y-axis based on reasonable total phosphorus values

example_plot
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r}

#Ensure 'month' is treated as a factor 
peter_paul_data$month <- factor(peter_paul_data$month,
                                levels = 1:12,
                                labels = month.abb)

# (a) Boxplot for Temperature by Month and Lake
plot_temp <- ggplot(peter_paul_data, aes(x = month, y = temperature_C, color = lakename)) +
  geom_boxplot() +
  labs(title = "Temperature by Month", y = "Temperature (°C)", x = "Month") +
  theme(
    legend.position = "none",  # Remove legend for this plot
    axis.title.x = element_blank(),  # Remove x-axis title for alignment
    axis.title.y = element_blank()  # Remove x-axis title for alignment
  )+
  ylim(0,30)

# (b) Boxplot for Total Phosphorus (TP) by Month and Lake
plot_tp <- ggplot(peter_paul_data, aes(x = month, y = tp_ug, color = lakename)) +
  geom_boxplot() +
  labs(title = "Total Phosphorus by Month", y = "TP (µg/L)", x = "Month") +
  theme(
    legend.position = "none",  # Remove legend for this plot
    axis.title.x = element_blank(),  # Remove x-axis title for alignment
    axis.title.y = element_blank()  # Remove x-axis title for alignment
  )+
  ylim(0,160)

# (c) Boxplot for Total Nitrogen (TN) by Month and Lake
plot_tn <- ggplot(peter_paul_data, aes(x = month, y = tn_ug, color = lakename)) +
  geom_boxplot() +
  labs(title = "Total Nitrogen by Month", y = "TN (µg/L)", x = "Month") +
  theme(
    legend.position = "none",  # Remove legend for this plot
    axis.title.x = element_blank(),  # Remove x-axis title for alignment
    axis.title.y = element_blank()  # Remove x-axis title for alignment
  )+
  ylim(0,3500)

plot_temp
plot_tp
plot_tn

```


```{r}
# Combine the plots into one grid while aligning axes
combined_plot <- plot_grid(
  plot_temp, 
  plot_tp, 
  plot_tn, 
  ncol = 3,                     # Arrange plots in 1 row with 3 columns
  align = "hv"                  # Align plots horizontally and vertically
)

# Add a shared legend from one of the plots (e.g., from plot_temp)
legend <- get_legend(ggplot(peter_paul_data, aes(x = month, color = lakename)) + 
                        geom_boxplot() + 
                        theme(legend.position = "bottom") + 
                        labs(title = "Lake Name") + 
                        theme_minimal() +
                        theme(
                           axis.title.x = element_blank(),  # Remove x-axis title for alignment
                        )+ labs(color = "Lake Name"))

# Combine the plots and the legend into the final layout
final_plot <- plot_grid(combined_plot, legend, ncol = 1, rel_heights = c(0.5, 0.1)) # Adjust legend size if needed

# Display the final combined plot
print(final_plot)
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature: Peaks in summer (July/August) with higher variability. Paul Lake generally shows slightly higher temperatures than Peter Lake. Total Phosphorus & Nitrogen: Consistent medians, more variability in warmer months. Paul Lake tends to have higher Total Phosphorus levels, while Peter Lake shows higher Total Nitrogen levels during certain months.
Lakes Comparison: Similar seasonal trends, but Paul Lake has higher temperatures and Total Phosphorus, whereas Peter Lake has higher Total Nitrogen in some months.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

```{r}

# Filter the dataset for the "Needles" functional group
needles_data <- niwo_litter_data %>% filter(functionalGroup == "Needles")

# Plot the data
ggplot(needles_data, aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  labs(title = "Dry Mass of Needle Litter by Date",
       x = "Date",
       y = "Dry Mass (g)",
       color = "NLCD Class") 

#By date or daynum?

```


7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}

# Plot the data with facets

ggplot(needles_data, aes(x = collectDate, y = dryMass)) +
  geom_point() +
  facet_wrap(~nlcdClass)+
  labs(title = "Dry Mass of Needle Litter by Date",
       x = "Date",
       y = "Dry Mass (g)") 

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I would lean towards Plot 6 for its potential to clearly show seasonal dry mass trends and differences between NLCD classes, which can be very insightful for environmental analysis.